---
title: "COVID-19 Report"
author: "Robert Forrest"
date: "April 11th 2024"
output: pdf_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Intro
In this report, I'll be taking at look at US COVID-19 data from John Hopkins University.
The data can be found hosted on their github here: https://github.com/CSSEGISandData/COVID-19.
In it contains the number of cases and deaths for each date for all states in the US.
I will perform some tidying of the data, data visualizations with analysis, create
a model and go through some conclusions. I am interested in exploring how the 
cases and deaths trended over time for the US as well as how deadly the virus was.


## Import Data
Read in 2 csv files from John Hopkin's github repo.
```{r importing}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv", 
                "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)

# read data in
us_cases <- read_csv(urls[1], show_col_types = FALSE)
us_deaths <- read_csv(urls[2], show_col_types = FALSE)
```


## Tidy and Transform
To clean the data, I will transform the data from a wide format into a long format.
This will allow for better analysis later on. I will also drop some unnecessary
columns and create some new columns.
``` {r tidy1}

us_cases_by_state <- us_cases %>%
	pivot_longer(cols = -c(UID, iso2, iso3, code3, FIPS, Admin2, Province_State,
	                       Country_Region, Lat, Long_, Combined_Key),
							names_to = "date",
							values_to = "cases") %>%
  mutate(date = mdy(date)) %>%
  select(-c(UID, iso2, iso3, code3, FIPS, Admin2, Country_Region, Lat, Long_,
            Combined_Key)) %>%
  group_by(Province_State, date) %>%
  summarize(cases = sum(cases), .groups = 'keep')


us_deaths_by_state <- us_deaths %>%
	pivot_longer(cols = -c(UID, iso2, iso3, code3, FIPS, Admin2, Province_State,
	                       Country_Region, Lat, Long_, Combined_Key, Population),
							names_to = "date",
							values_to = "deaths") %>%
  mutate(date = mdy(date)) %>%
  select(-c(UID, iso2, iso3, code3, FIPS, Admin2, Country_Region, Lat, Long_,
            Combined_Key, Population)) %>%
  group_by(Province_State, date) %>%
  summarize(deaths = sum(deaths), .groups = 'keep')
```

Here I will combine the data for cases and deaths into a single table. I will
also add a deaths_per_cases column.
``` {r tidy2}
us_all_data <- us_cases_by_state %>%
  full_join(us_deaths_by_state, by=c("Province_State", "date")) %>%
  mutate(deaths_per_cases = deaths / cases)
```

Its important to get rid of any potentially bad data. Here I will make sure that
the deaths_per_cases do not exceed 1.
``` {r tidy3}
us_all_data <- us_all_data %>%
  filter(deaths_per_cases < 1) %>%
  arrange(Province_State, date)
```


## Visualization and Analysis
Lets take a look at the top 5 states with the most cases and deaths.
``` {r vis1}
cases_and_deaths_by_state <- us_all_data %>%
  group_by(Province_State) %>%
  summarise(total_cases = sum(cases), total_deaths = sum(deaths)) %>%
  arrange(desc(total_cases))

# This one is sorted by cases
head(cases_and_deaths_by_state, 5)

# This one is sorted by deaths
head(cases_and_deaths_by_state %>% arrange(desc(total_deaths)), 5)

```
From the results, California had the most cases and deaths. Lets take a look
at how California's cases and deaths trended over time.

``` {r vis2}
ggplot(data = subset(us_all_data, Province_State == "California"),
       aes(x = date, y = cases)) +
  geom_line() +
  labs(x = "Date", y = "Cases", title = "COVID-19 Cases in California Over Time")
```
Its interesting to see that were appears to be 2 major spikes in the number of
cases near the end of 2020 and the beginning of 2022. I believe this could be
due to the heavy amount of holiday traveling going on around those times.

``` {r vis3}
ggplot(data = subset(us_all_data, Province_State == "California"),
       aes(x = date, y = deaths)) +
  geom_line() +
  labs(x = "Date", y = "Deaths", title = "COVID-19 Deaths in California Over Time")
```
When taking at look the deaths, it makes sense that given the large amount of
cases near 2021, that there would also be a spike in the deaths. The vaccine was
not widely available in time to prevent the deaths. When comparing the second
spike in cases in 2022, we don't see such a drastic increase in deaths, which
is most likely due to people either having the vaccine already or they may have
already gotten sick prior.

Here we will take a look at the number of deaths per cases over time. This will
allow us to understand how deadly the virus was in California.
``` {r vis4}
ggplot(data = subset(us_all_data, Province_State == "California"),
       aes(x = date, y = deaths_per_cases)) +
  geom_line() +
  labs(x = "Date", y = "Deaths per Cases", title = "COVID-19 Deaths per Cases in California Over Time")
```

Next, lets took at the US as a whole to understand how the deaths per cases is
trending over time.
``` {r vis5}
us_cases_and_deaths_by_date <- us_all_data %>%
  group_by(date) %>%
  summarise(total_cases = sum(cases), total_deaths = sum(deaths)) %>%
  mutate(deaths_per_cases = total_deaths / total_cases) %>%
  filter(deaths_per_cases < 1)

ggplot(us_cases_and_deaths_by_date,
       aes(x = date, y = deaths_per_cases)) +
  geom_line() +
  labs(x = "Date", y = "Deaths per Cases", title = "COVID-19 Deaths per Cases in the US Over Time")
```
It looks like when using at the entire United States, the trend is similar
to what happened in California. The virus appears very deadly in the beginning
but sometime after 2022, it begins to consistently stay lower.

Finally, lets see if the number of cases have any relationship with the number
of deaths. A model will be created for it later.
``` {r vis6}
ggplot(us_cases_and_deaths_by_date,
       aes(x = total_cases, y = total_deaths)) +
  geom_line() +
  labs(x = "Total Cases", y = "Total Deaths", title = "COVID-19 Deaths vs Cases in the US")
```
It looks like there is a relationship. So as the number of cases increases, the
number of deaths do as well.


## Model
Now a model will be created that will attempt to predict the total number of
deaths, based on the total number of cases.
``` {r model1}
mod <- lm(total_deaths ~ total_cases, data= us_cases_and_deaths_by_date)

summary(mod)
```
The results from the linear model look good. The p-value is < 0.05 and the 
R-squared is close to 1.

Now we can visualize the actual vs predicted deaths.
``` {r model2}
us_cases_and_deaths_by_date <- us_cases_and_deaths_by_date %>% mutate(pred_deaths=predict(mod))

us_cases_and_deaths_by_date %>%
  ggplot() +
  geom_point(aes(x = total_cases, y = total_deaths, color = "Actual")) +
  geom_point(aes(x = total_cases, y = pred_deaths, color = "Predicted")) +
  labs(title = "Actual vs Predicted Deaths",
       x = "Total Cases",
       y = "Total Deaths") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red"),
                     labels = c("Actual", "Predicted")) +
  guides(color = guide_legend(title = "Legend"))

```
The model appears to fit well enough, but isn't very good at the extreme ends
of the range or the middle. This is likely cause by external factors such as
vaccines since the cases can increase drastically without leading to more deaths.


## Conclusions and Bias Identification
In conclusion, COVID-19 started off a very deadly disease but as time went
on and we learned more about its behavior, we were able to reduce its deadliness.
Vaccines were also a likely contributor to the drastic decrease in deaths and
cases. We did not experience an increase in deaths or cases near the beginning
of 2023 like what was experienced for the past 2 years.

One source of bias in this data would be in how the data was gathered. The cases
would be under represented by some amount since not everyone will report that they
got the virus and would have likely just stayed at home. The number of deaths
have a better chance of being more correct since any cause of death would eventually
be reported. 

In my analysis, I also only looked at the US since that is where I am based. The
virus could have a wildly different behavior or outcome if it was in another
country. John Hopkins does have more data on COVID-19 from other countries available.
There could also be a misrepresentation of cases and deaths in this data since
countries have different reporting practices.

Something else to keep in mind when thinking about bias, is that more populated
areas would naturally have more cases and deaths. Data that is aggregated at the 
state level, may not be representative for someone in a less populated location of
that same state.

## Show Session Info
``` {r info}
sessionInfo()
```
