---
title: "NYPD Shooting Incident Report"
author: "Robert Forrest"
date: "April 11th 2024"
output: pdf_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```
## Intro

In this report I will import and clean the NYPD shooting incident data and then 
do some analysis. Afterwards I will create a model that will predict the number
of shootings expected.


## Import Data
Reads in the data from the government dataset catalog website.
```{r importing}
data <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv",
                 show_col_types = FALSE)
summary(data)
```


## Tidy and Transform
To clean the data, I will drop columns that I'm not interested in and will fix
the occur_date column by transforming it into a real date column.
```{r tidy1}
data <- data %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  select(-c(INCIDENT_KEY, JURISDICTION_CODE, LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC,
            X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat))
print(names(data))
```


Look at unique values in order to do some potential replacements.
``` {r tidy2}
# These are the only columns I want to look at for unique values.
columns_to_check <- setdiff(names(data), c("OCCUR_DATE", "OCCUR_TIME", "PRECINCT"))

# Custom function to display counts of unique values in each column
print_unique_counts <- function(df) {
  for (col in names(df)) {
    cat("Column:", col, "\n")
    cat("------------\n")
    counts <- table(df[[col]])
    for (val in names(counts)) {
      cat(val, ": ", counts[val], "\n")
    }
    cat("\n")
  }
}

print_unique_counts(data[columns_to_check])
```


Next I will replace empty string data or unknown-like/messy data with "UNKNOWN".
```{r tidy3}
# Define function to handle replacement unknown-like values
replace_with_unknown <- function(x) {
  ifelse(x == "" | x == "(null)" | is.na(x) | x == "U" | x == "1022" | x == "1020"
         | x == "940" | x == "224", "UNKNOWN", x)
}

data <- data %>%
  mutate(across(all_of(columns_to_check), ~ replace_with_unknown(.)))

print_unique_counts(data[columns_to_check])
```


Next I'll add some more columns of interest.
``` {r tidy4}
data <- data %>%
  mutate(DAY_OF_WEEK = wday(OCCUR_DATE, label = TRUE, abbr = FALSE)) %>%
  mutate(DAYS_SINCE_DATA_START = as.numeric(OCCUR_DATE - min(data$OCCUR_DATE)))

data_start_date = format(min(data$OCCUR_DATE), "%Y-%m-%d")

cat("Data start date=", data_start_date)
```


## Visualization and Analysis
This is a plot of the number of shootings by Borough.
``` {r vis1}
event_counts_boro <- data %>%
  group_by(BORO) %>%
  summarise(NUM_SHOOTINGS = n())

ggplot(event_counts_boro, aes(x = BORO, y = NUM_SHOOTINGS)) +
  geom_bar(stat = "identity", fill = "lightblue", color = "black") +
  labs(title = "Number of Shootings by Borough",
       x = "Borough",
       y = "Number of Shootings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
So here we see that Brooklyn has the highest number of shootings and Staten Island
has the lowest. Its possible that the differences are related to the population,
so further analysis would be required in order to rule that possibility out.


This is a plot of the number of shootings by day of the week.
``` {r vis2}
event_counts_day <- data %>%
  group_by(DAY_OF_WEEK) %>%
  summarise(NUM_SHOOTINGS = n())

ggplot(event_counts_day, aes(x = DAY_OF_WEEK, y = NUM_SHOOTINGS)) +
  geom_bar(stat = "identity", fill = "lightgreen", color = "black") +
  labs(title = "Number of Shootings by Day of the Week",
       x = "Day of the Week",
       y = "Number of Shootings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
It looks like the weekends are when shootings happen the most. I suspect that its
due to people being more "free" during the weekends typically. Or if these are
happening at night, people are more likely to me active during weekend nights vs
weekday nights.


Plot the cumulative number of events over time.
``` {r vis3}
event_counts_cumu <- data %>%
  group_by(OCCUR_DATE) %>%
  summarise(NUM_SHOOTINGS = n()) %>%
  arrange(OCCUR_DATE) %>%
  mutate(CUMULATIVE_SHOOTINGS = cumsum(NUM_SHOOTINGS))

ggplot(event_counts_cumu, aes(x = OCCUR_DATE, y = CUMULATIVE_SHOOTINGS)) +
  geom_line(color = "blue") +
  labs(title = "Cumulative Number of Shootings Over Time",
       x = "Date",
       y = "Total Shootings") +
  theme_minimal()
```

Based on the plot above, it looks like the shootings are going up in a straight line.
This makes me wonder if there are certain age groups that are more at risk. So
here I will plot a heatmap to understand which groups are involved with the most
shootings.

``` {r analysis1}
# Filter out rows where either PERP_AGE_GROUP or VIC_AGE_GROUP is "UNKNOWN"
filtered_data <- data %>%
  filter(PERP_AGE_GROUP != "UNKNOWN" & VIC_AGE_GROUP != "UNKNOWN")

# Create a table of counts for each combination of PERP_AGE_GROUP and VIC_AGE_GROUP
heatmap_data <- table(filtered_data$PERP_AGE_GROUP, filtered_data$VIC_AGE_GROUP)

# Add row and column sums for cumulative counts
heatmap_data <- addmargins(heatmap_data)

# Plot the heatmap with counts within each cell
ggplot(as.data.frame(heatmap_data), aes(x = Var2, y = Var1, fill = Freq, label = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(color = "black") +
  labs(title = "Shootings Heatmap",
       x = "Victim Age Group",
       y = "Perpetrator Age Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
```
These results tell us that the perpetrator group causing the most shootings are
the 18-24 year olds. They mostly go after other 18-24 year olds. The group most
victimized are the 25-44 year olds. They are mostly victimized by other 25-44 year
olds. Contrary to what one might think, the older people (45+) are not victims as often
as younger people.


## Model
Create a model to predict the total number of shootings, based on the data start date.
``` {r model1}
event_counts_for_model <- data %>%
  group_by(DAYS_SINCE_DATA_START, BORO, PRECINCT, VIC_AGE_GROUP, PERP_AGE_GROUP) %>%
  summarise(NUM_SHOOTINGS = n(), .groups = "drop") %>%
  arrange(DAYS_SINCE_DATA_START) %>%
  mutate(CUMULATIVE_SHOOTINGS = cumsum(NUM_SHOOTINGS))


mod <- lm(CUMULATIVE_SHOOTINGS ~ DAYS_SINCE_DATA_START, data= event_counts_for_model %>%
 filter(PERP_AGE_GROUP != "UNKNOWN" & VIC_AGE_GROUP != "UNKNOWN"))

summary(mod)
```
The results from the linear model look good. The p-value is < 0.05 and the 
R-squared is close to 1.


Visualize the predicted shootings vs actual shootings.
``` {r model2}
# Create a data frame for predictions with the same structure as event_counts_for_model
new_data <- data.frame(DAYS_SINCE_DATA_START = event_counts_for_model$DAYS_SINCE_DATA_START)

# Add a column for the predicted values
event_counts_for_model$predicted <- predict(mod, newdata = new_data)

# Now can plot and see actual vs predicted
event_counts_for_model %>% 
  ggplot() +
  geom_point(aes(x = DAYS_SINCE_DATA_START, y = CUMULATIVE_SHOOTINGS, color = "Actual")) +
  geom_point(aes(x = DAYS_SINCE_DATA_START, y = predicted, color = "Predicted")) +
  labs(title = "Actual vs Predicted Shootings",
       x = paste("Days since", data_start_date),
       y = "Cumulative Shootings") +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red"),
                     labels = c("Actual", "Predicted")) +
  guides(color = guide_legend(title = "Legend"))
```


## Conclusions and Bias Identification
In conclusion, there are areas in New York that are more dangerous than others.
The primary victims and perpetrators are from 2 different age groups, but are
still rather close in age. There unfortunately doesn't appear to be a noticeable
decrease in shootings over time.

A source of bias that I have when analyzing this data is in how I handled missing
or unknown values. I decided to make all missing and unknown values equal to "UNKNOWN".
This means that my analysis and model could greatly under represent reality if it
turns out that the most important shootings are originating from the "UNKNOWN"
category.

A source of bias from this data, would be due to the shootings only originating
from New York. There is almost a universal mindset that New York can be particularly
dangerous. Therefore, its important to keep in mind that the data collected here
and the associated model, should not be used for other cities. Also, this only
has data reported from NYPD. There could be many shootings happening that go
unreported or that are in a different police district.

Due to these biases, its important to mitigate them by having the audience
understand that they exist and to gather data from other sources before making
further conclusions.

## Show Session Info
``` {r info}
sessionInfo()
```